% -*- coding: utf-8 -*-

\begin{zhaiyao}
\begin{spacing}{1.5}
{

水下场景的新视角合成是海洋科学观测、水下机器人导航、海底资源勘探与生态监测中的基础视觉任务。与陆地场景相比，水体介质会引入波长相关的光学衰减和后向散射，使采集图像出现蓝绿色偏、对比度下降和细节模糊；同时，水流扰动、平台抖动以及鱼类、珊瑚等非刚性目标的运动又会破坏静态多视角一致性假设。现有静态水下重建方法能够显式处理水体退化，但难以建模动态扰动；通用动态新视角合成方法虽可表示时变几何，却通常忽略水下成像物理，容易产生颜色失真和高斯漂浮伪影。针对上述问题，本文以4DGaussians为动态场景表示骨架，将SeaSplat中的水下图像形成模型迁移至动态高斯渲染流程，构建面向扰动水下环境的新视角合成系统。系统在光栅化后引入后向散射网络和直接透射衰减网络，利用渲染深度估计水体介质对颜色的影响；同时引入Depth Anything V2生成的单目逆深度先验，通过尺度归一化的深度一致性损失约束高斯体空间分布；在训练策略上，本文采用从几何初始化、水下参数预热到联合优化的分阶段流程，以降低水下网络、动态形变网络与高斯属性之间的梯度干扰。实验在Robot、Fish、Coral和Streaks四个动态水下场景以及SeaThru-NeRF静态水下数据集上进行。动态场景实验中，本文方法在Robot、Coral和Streaks三个场景上取得最优PSNR，平均PSNR达到30.54 dB，较UDR-GS平均提升约1.6\%，较4DGS平均提升约3.1\%；在Fish场景中，本文方法的PSNR略低于UDR-GS，表明高速非刚性运动仍会给物理颜色恢复与动态形变联合优化带来挑战。消融实验进一步表明，水下物理建模和单目深度监督在不同场景中的贡献并不完全相同：深度监督对几何稳定性提升更为直接，水下物理模型在色偏和散射较明显且运动相对稳定的场景中收益更明显。

}
\end{spacing}
\end{zhaiyao}


\begin{guanjianci}
三维高斯溅射；水下场景重建；动态场景建模；水下物理成像；深度监督
\end{guanjianci}


\begin{abstract}
\begin{spacing}{1.5}
Underwater novel view synthesis is a fundamental vision task for marine observation, underwater robot navigation, seafloor inspection, and ecological monitoring. Compared with in-air scenes, underwater imagery suffers from wavelength-dependent attenuation and backscatter, which cause color shift, low contrast, and blurred details. Meanwhile, water flow, camera perturbation, moving fish, and deformable corals break the static multi-view consistency assumption. Existing static underwater reconstruction methods model medium degradation but cannot handle dynamic objects, whereas general dynamic reconstruction methods usually ignore underwater image formation and may produce color distortion and floating Gaussian artifacts.

This thesis builds a dynamic underwater novel view synthesis system by migrating the physically grounded image formation model of SeaSplat into the 4DGaussians framework. After differentiable Gaussian rasterization, a backscatter network and a direct attenuation network estimate depth-dependent medium effects from rendered depth. In addition, monocular inverse-depth priors generated by Depth Anything V2 are used as weak supervision through a scale-normalized depth consistency loss. A staged training strategy is adopted to reduce gradient interference among Gaussian geometry, dynamic deformation, and underwater medium parameters.

Experiments are conducted on four dynamic underwater scenes, namely Robot, Fish, Coral, and Streaks, as well as static scenes from the SeaThru-NeRF dataset. On the dynamic benchmark, the proposed method obtains the best PSNR on Robot, Coral, and Streaks, and reaches an average PSNR of 30.54 dB, improving over UDR-GS by about 1.6\% and over 4DGS by about 3.1\%. On Fish, the PSNR is slightly lower than UDR-GS, indicating that fast non-rigid motion remains challenging for joint optimization of dynamic deformation and physical color restoration. Ablation studies show that monocular depth supervision directly improves geometric stability, while the underwater physical model is most beneficial when color shift and scattering are evident and motion is relatively stable.
\end{spacing}
\end{abstract}


\begin{keywords}
3D Gaussian splatting; underwater scene reconstruction; dynamic scene modeling; underwater physical imaging; depth supervision
\end{keywords}
